---
output:
  md_document:
    variant: markdown_github
---

# Purpose

The purpose of this folder is to practice using an application programming interface for my econometrics tutorial.


```{r}

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
if (!require("pacman", quietly = TRUE )) {
    install.packages("pacman")
    library(pacman)
}
if (!require("tibble", quietly = TRUE )) {
    install.packages("tibble")
    library(tibble)
}

if (!require("knitr", quietly = TRUE )) {
    install.packages("knitr")
    library(knitr)
}

if (!require("kableExtra", quietly = TRUE )) {
    install.packages("kableExtra")
    library(kableExtra)
}

p_load(wbstats,WDI,tidyverse,haven)

list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```

# Setup

```{r}
# Verify that the packages are ready
library(help="WDI") # lists functions in the WDI package
wbstats::wb_cachelist # glimpse of cached World Bank metadata
WDI::WDI_data # explore WDI data structures
```


# Finding Indicator Codes

Each World Banking indicator has a unique alphanumeric code (e.g., NY.GDP.PCAP.KD). The code is the key that the API uses to retrieve the correct seires. Both wbstats and WDI provide search function.

## Usng *wbstats* (Recommended)

```{r}
# Search for indicators containing "GDP"
gdp_indicators <- wb_search("GDP")
head(gdp_indicators)

# Narrow search to "GDP per capita "
gdp_pc <- wb_search("GDP per capita")
View(gdp_pc) #interactive table viewer in RStudio

# Important columns :
 # indicator _id -> code for wb_ data ()
 # indicator -> human - readable description
 # unit -> measurement unit (e.g., " constant 2015 US$")
```

## Using *WDI* (Classic)

```{r}
# Search for indicators containing " inflation "
WDIsearch("inflation")
```

# Understanding Indicator Code Suffixes

The suffix of an indicator code encodes crucial information about the unit, adjustment method, and base year. Misinterpreting suffixes is a common source of errors in empirical work.

```{r}
wb_suffixes <- tribble(
  ~Suffix, ~`Meaning and Appropriate Use`,
  ".CD", "Current USD (nominal). Affected by both inflation and exchange rate movements. Use for headline figures, single-period snapshots, or when nominal values are explicitly required.",
  ".KD", "Constant USD (real, inflation-adjusted). Base year is usually 2015. Use for comparing growth rates over time within a country.",
  ".PP.CD", "Purchasing Power Parity (PPP) adjusted, expressed in current international dollars. Corrects for cross-country differences in price levels. Use for comparing living standards across countries at a point in time.",
  ".PP.KD", "PPP-adjusted, expressed in constant international dollars. Combines inflation adjustment with PPP conversion. Use for comparing growth in living standards across countries over time.",
  ".ZS", "Percentage share (ratio). Already unit-free (e.g., trade as % of GDP, unemployment rate). Values are typically between 0 and 100.",
  ".GD", "Share of GDP. Similar to .ZS but explicitly relative to GDP.",
  ".LN", "Natural logarithm of the underlying series. Rarely used directly; often employed in internal calculations.",
  ".MA", "Moving average or smoothed series. Indicates that the value has been smoothed over time.",
  ".DT", "Decadal data (e.g., population census every 10 years). Check the indicator description for the exact frequency.",
  ".AR", "Annual Report data. May differ slightly from standard series due to reporting conventions.",
  ".EST", "Estimated value (not directly reported by the country). Used by World Bank staff to fill gaps."
)

# Generate the formatted table
wb_suffixes %>%
  kable(
    caption = "Table 1: Comprehensive List of Common World Bank Indicator Suffixes",
    booktabs = TRUE,
    col.names = c("Suffix", "Meaning and Appropriate Use")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  ) %>%
  column_spec(1, bold = TRUE, border_right = TRUE) %>%
  column_spec(2, width = "30em")
```

**Economist's rule of thumb**:

* Coss-Country level comparison -> use .PP.CD or .PP.KD (PPP-adjusted).

* Growth over time within a country -> use .KD (constant/real).

* Policy reports in nominal USD -> .CD (current/nominal).

# Country Codes and Time Granularity

## ISO Country Code: ISO2 vs ISO3

The WB API accepts both two-letter and three-letter country codes:

* **ISO2**: Often used in plots and quick identification. 

* **ISO3**: More informative and are the standard for merging with other international datasets(such as UN, IMF, or Penn Word Tables)

## Time Granularity in API Calls

The wbstats package automatically handles the format, but when constructing raw API calls, you must specify dates correctly:

* **Annual data**: Use YYYY

* **Quarterly data**: Use YYYYQ (2020Q1)

* **Monthly data**: Use YYYM (2020M01)

# Downloading Data: Core Workflow

## Using wbstats

```{r}
# Define indicators using a named vector ( names become column names )
my_indicators <- c(
gdp_per_capita = "NY.GDP.PCAP.KD", # constant 2015 USD
inflation = "FP.CPI.TOTL.ZG", # annual CPI inflation (%)
trade_openness = "NE.TRD.GNFS.ZS", # trade as % of GDP
school_enroll = "SE.SEC.ENRR" # secondary gross enrolment ratio
)

# Specify countries ( ISO2 or ISO3 codes )
 my_countries <- c("ZA","NG","KE","GH","EG")

 # Download data for 2000 -2023
africa_data <- wb_data(indicator = my_indicators, country = my_countries, start_date = 2000, end_date = 2023)

 # Inspect the structure
 head(africa_data)
 glimpse(africa_data) # displays column types and units
```

## Using WDI (Classic)

```{r}
wdi_data <- WDI(indicator = c(gdp_per_capita = "NY.GDP.PCAP.KD", inflation = "FP.CPI.TOTL.ZG"), country = c("ZA","NG","KE","GH","EG"), start = 2000, end = 2023)

head(wdi_data)
```

