---
output:
  md_document:
    variant: markdown_github
---

# Purpose

The purpose of this folder is to practice using an application programming interface for my econometrics tutorial.

```{r}

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
if (!require("pacman", quietly = TRUE )) {
    install.packages("pacman")
    library(pacman)
}
if (!require("tibble", quietly = TRUE )) {
    install.packages("tibble")
    library(tibble)
}

if (!require("knitr", quietly = TRUE )) {
    install.packages("knitr")
    library(knitr)
}

p_load(wbstats,WDI,tidyverse,haven)

list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```

# Setup

```{r}
# Verify that the packages are ready (UNCOMMENT IF YOU WANT TO)
#library(help="WDI") # lists functions in the WDI package
#wbstats::wb_cachelist # glimpse of cached World Bank metadata
#WDI::WDI_data # explore WDI data structures
```

# Finding Indicator Codes

Each World Banking indicator has a unique alphanumeric code (e.g., NY.GDP.PCAP.KD). The code is the key that the API uses to retrieve the correct seires. Both wbstats and WDI provide search function.

## Usng *wbstats* (Recommended)

```{r}
# Search for indicators containing "GDP"
gdp_indicators <- wb_search("GDP")
head(gdp_indicators)

# Narrow search to "GDP per capita "
gdp_pc <- wb_search("GDP per capita")
head(gdp_pc) #interactive table viewer in RStudio

# Important columns :
 # indicator _id -> code for wb_ data ()
 # indicator -> human - readable description
 # unit -> measurement unit (e.g., " constant 2015 US$")
```

## Using *WDI* (Classic)

```{r}
# Search for indicators containing " inflation "
WDIsearch("inflation")
```

# Understanding Indicator Code Suffixes

The suffix of an indicator code encodes crucial information about the unit, adjustment method, and base year. Misinterpreting suffixes is a common source of errors in empirical work.

```{r}
# Create the data
wb_data <- tribble(
  ~Suffix, ~`Meaning and Appropriate Use`,
  ".CD", "Current USD (nominal). Affected by both inflation and exchange rate movements.",
  ".KD", "Constant USD (real, inflation-adjusted). Base year is usually 2015.",
  ".PP.CD", "PPP-adjusted, current international dollars. Corrects for cross-country price levels.",
  ".PP.KD", "PPP-adjusted, constant international dollars. Combines inflation adjustment with PPP.",
  ".ZS", "Percentage share (ratio). Already unit-free (e.g., trade as % of GDP).",
  ".GD", "Share of GDP. Similar to .ZS but explicitly relative to GDP.",
  ".LN", "Natural logarithm of the underlying series. Used in internal calculations.",
  ".MA", "Moving average or smoothed series. Indicates values have been smoothed.",
  ".DT", "Decadal data (e.g., population census every 10 years).",
  ".AR", "Annual Report data. May differ from standard series due to reporting.",
  ".EST", "Estimated value (not directly reported). Used to fill gaps."
)

# Generate a Markdown-compatible table
kable(wb_data, format = "markdown", caption = "Table 1: World Bank Indicator Suffixes")
```

**Economist's rule of thumb**:

- Coss-Country level comparison -\> use .PP.CD or .PP.KD (PPP-adjusted).

- Growth over time within a country -\> use .KD (constant/real).

- Policy reports in nominal USD -\> .CD (current/nominal).

# Country Codes and Time Granularity

## ISO Country Code: ISO2 vs ISO3

The WB API accepts both two-letter and three-letter country codes:

- **ISO2**: Often used in plots and quick identification.

- **ISO3**: More informative and are the standard for merging with other international datasets(such as UN, IMF, or Penn Word Tables)

## Time Granularity in API Calls

The wbstats package automatically handles the format, but when constructing raw API calls, you must specify dates correctly:

- **Annual data**: Use YYYY

- **Quarterly data**: Use YYYYQ (2020Q1)

- **Monthly data**: Use YYYM (2020M01)

# Downloading Data: Core Workflow

## Using wbstats

```{r}
# Define indicators using a named vector ( names become column names )
my_indicators <- c(
gdp_per_capita = "NY.GDP.PCAP.KD", # constant 2015 USD
inflation = "FP.CPI.TOTL.ZG", # annual CPI inflation (%)
trade_openness = "NE.TRD.GNFS.ZS", # trade as % of GDP
school_enroll = "SE.SEC.ENRR" # secondary gross enrolment ratio
)

# Specify countries ( ISO2 or ISO3 codes )
 my_countries <- c("ZA","NG","KE","GH","EG")

 # Download data for 2000 -2023
africa_data <- wb_data(indicator = my_indicators, country = my_countries, start_date = 2000, end_date = 2023)

 # Inspect the structure
 head(africa_data)
 glimpse(africa_data) # displays column types and units
```

## Filtering by Region or Income Group

```{r}
# Retrieve country metadata
country_info <- wb_countries()

# Extract ISO3 codes for Sub - Saharan Africa
ssa_countries <- country_info %>%
filter(region == "Sub-Saharan Africa") %>%
pull(iso3c)

# Download GDP per capita for all SSA countries
ssa_data <- wb_data (indicator = c(gdp_pc = "NY.GDP.PCAP.KD"), country = ssa_countries, start_date = 2000, end_date = 2022)
# Approximately 48 countries 23 years = ~1 ,104 observations
```

## Understanding The Returned Dataframe

wbstats returns a tidy dataframe where each row corresponds to one country-year observation. Key columns include:

```{r}
# Summarise missing values across all columns
africa_data %>% summarise(across(everything(), ~ sum(is.na(.))))

# Note : Missing data often reflects non - reporting , which may correlate with
# institutional quality . Consider the implications for your analysis .
```
